Avoiding tipping points in fisheries management through Gaussian Process Dynamic Programming

### 2014-11-07 manuscripts/manuscript.Rmd
update data file names to match figures

Decision making under uncertainty is a ubiquitous challenge in the
management of human intervention in natural resources and conservation.
Ecological dynamics are frequently complex and difficult to measure,
making uncertainty in our understanding a persistent challenge.
Decision-theoretic approaches provide a framework to determine the
best sequence of actions in face of uncertainty, but only when that
uncertainty can be meaningfully quantified [@Fischer2009].  Over the last
four decades (but stretching even further back in time), marked by  the
seminal contributions of @Clark1976, @Clark2009 and @Walters1978,
dynamic optimization methods, particularly Stochastic Dynamic Programming
(SDP), have become increasingly important as a means of
understanding how to manage human intervention into natural systems.
Simultaneously, there has been increasing recognition of the importance of
Allee population sizes or `tipping points' [@Scheffer2001;
@Polasky2011] that management actions may inadvertently
cause populations to cross.  @Scheffer2009 emphasized the difficulties of
formulating even qualitatively correct models of the underlying processes.

We develop a novel approach to address these concerns in the context of
fisheries; though the underlying challenges and methods are germane to
many other problems of conservation or natural resource exploitation,
such as forestry.  Economic value and ecological concern have made
marine fisheries the crucible for much of the founding work [@Gordon1954;
@Clark1976; -@Clark2009; @Reed1979; @May1979; @Ludwig1982] on management
under uncertainty.  Global declines [@Worm2006] and the controversy
surrounding their interpretation [@Hilborn2007; @Worm2009] make
understanding these challenges all the more pressing.

@@

Uncertainty enters the decision-making process at many levels:
intrinsic stochasticity in biological processes, measurements, and
implementation of policy [_e.g._ @Reed1979; @Clark1986; @Roughgarden1996;
@Sethi2005], parametric uncertainty [_e.g._ @Ludwig1982; @Hilborn1997;
@McAllister1998; @Schapaugh2013], and model or structural uncertainty
[_e.g._ @Williams2001; @Cressie2009;  @Athanassoglou2012].  Of these,
structural uncertainty is generally the hardest to quantify. Typical approaches
assume a weak notion of model uncertainty in which the correct model
(or reasonable approximation) of the dynamics must be identified from
among a handful of alternative models, using either model choice, model averaging, or 
introducing yet greater model complexity of which others may be special
cases (model averaging being one such way to construct such a model)
[@Williams2001; @Athanassoglou2012; @Cressie2009].  Even setting aside
other computational and statistical concerns (e.g. @Cressie2009), these
approaches do not address our second concern - representing uncertainty
outside the observed data range.

@@

Model uncertainty is particularly insidious when model predictions must
be made outside of the range of data on which the model was estimated, and models may be abused [@Mangel2001].
Extrapolation uncertainty is felt most keenly in decision-theoretic
(or optimal control) applications, since (a) exploring the space of 
all possible actions involves considering actions that may move the
system outside the range of observed behavior, and (b) decision-theoretic
algorithms rely not only on reasonable estimates of the expected
outcomes, but depend on the weights given to all possible outcomes
[_e.g._ @Weitzman2013], hence estimates of the uncertainty can be as
important as estimates of the average or expected outcome.

Concerns about tipping points in ecological dynamics [@Scheffer2001;
@Polasky2011] highlight this problem because they arise from this very
situation.  In such cases, precise models are not available and data is
limited or biased to a particular region of state space, such as high
stock levels or an otherwise desirable state for the ecosystem. If such
models or data were available, one would know just how far a system could
be pushed before going over the tipping point, and management would be
simple.  The fact that this concern is frequently discussed highlights
how often we have only limited data and inperfect models. Moreover,
the case of tipping points highlights how even small errors can have
very large consequences, as we shall illustrate later.

The dual concerns of model uncertainty and incomplete data coverage
pose a substantial challenge to existing decision-theoretic approaches
[@Brozovic2011].  Because intervention is often (but not always, see
@Hughes2013) too late after a tipping point has been crossed, management
is most often concerned with avoiding potentially catastrophic
tipping points before any data are available at or following a
transition that would more clearly reveal these regime shift dynamics
[e.g. @Bestelmeyer2012].

We illustrate how a Stochastic Dynamic Programming (SDP) algorithm
[@Mangel1988; @Marescot2013] can be driven by the predictions from a
Bayesian Non-Parametric (BNP) model of population dynamics [@Munch2005a].
This provides two distinct advantages compared with contemporary
approaches.  First, using a BNP sidesteps the need for an accurate
model-based description of the system dynamics.  Second, the BNP can
better reflect uncertainty that arises when extrapolating a model outside
of the data on which it was fit.  We illustrate that when the correct
model is not known, this latter feature is crucial to providing a robust
decision-theoretic approach in face of substantial structural uncertainty.
We use _robust_ to distinguish approaches that provide nearly optimal solutions
(as measured by the chosen value function) without being sensitive to the
choice of the (unknown) underlying model.

<!--
FIXME We need to be 
-->

This paper represents the first ecological application of the
SDP decision-making framework without an a priori model of the
underlying dynamics.  In contrast to parametric models which can only
reflect uncertainty in parameter estimates, the BNP approach provides a
state-space dependent representation of uncertainty.  This permits a much
greater uncertainty far from the observed data than near the observed
data.  These features allow the Gaussian Process Dynamic Programming
(GPDP) approach to find robust management solutions in face of limited
data and without knowledge of the correct model structure.

@@

To illustrate GPDP, we consider the performance of the GPDP policy against
the policies derived under several alternative parametric models [@Reed1979; @Ludwig1982; @Mangel1988]. 
The nature of decision-making problems provides a compelling and
pragmatic way to compare models: Rather than compare models in terms
of best fit to data, we define model performance in the concrete terms
of the decision-maker's objectives.

@@

[^1]: Note that we abbreviate Gaussian Process as GP, which refers only to 
the statistical model we use to approximate the population dynamics (as opposed
to a familiar parametric model such as the Ricker equation), while we use the
term GPDP, Gaussian Process Dynamic Programming, to refer to the _use_ of a
GP as the underlying process model when solving a Dynamic Program, as we illustrate
in this paper.  Hence we will refer to the models as: GP, Ricker, Allen, etc, while
we refer to the novel method we put forward here as GPDP.

@@

Dynamic optimization  requires characterizing the dynamics of a state
variable (or variables), a control action, and a value function.
In this paper, we consider only a single state variable, so that the
ideas can be described as simply as possible. That is, we focus on a
class of one-dimensional models of population dynamics as a 'best-case'
scenario for the parametric model performance relative to the GP.
This is a best-case scenario for the parametric models because the
underlying dynamics are always simulated from one of the parametric models
in the set, whereas in reality we never have the "true" model among our
candidates.  By choosing one-dimensional models with few parameters, we
limit the chance that poor performance will be due to our inability to
estimate parameters well, something that becomes a more severe problem
for higher-dimensional parametric models. Further, the parametric models
we consider are those most commonly used in modeling stock-recruitment
dynamics or to model sudden transitions between alternative stable states.

@@

We let $x(t)$ denote the size (numbers or biomass) of the focal
population at time $t$ and assume that in the absence of take its
dynamics are. (Throughout, we will use $X(t+1)$ with a capital $X$ as
a reminder that this is a stochastic, uncertainty quantity at time $t$,
while we use lowercase $x(t)$ as a reminder that the current population
size is treated as a given without error.  The latter assumption is
straight-forward to relax but will simplify the presentation here).

\begin{equation}
X(t+1) = Z(t) f(x(t), p) \label{eq1} 
\end{equation}

Where Z(t) is log-normally distributed process stochasticity [@Reed1979]
and $p$ is a vector of parameters to be estimated from the data.
We describe three choices for $f(X(t),p)$ in the next section.In this
simple model, the control action is a harvest or take, $h(t)$, measured
in the same units as $X$, at time $t$.  Thus, in the presence of take,
the population size on the right hand side of Eqn 1 is replaced by
$S(t) = X(t) - h(t)$.

To construct the value function, we consider a return when $X(t) = x $ and
harvest $h(t) = h$ denoted as the reward, $R(x, h)$. For example, if the
return is simply the harvest at time $t$, then $R(x, h) = \min(x, h)$. We
assume that future harvests are discounted relative to current ones at
a constant rate of discount $\delta$ and ask for the harvest policy
that maximizes total discounted harvest between the current time $t$
and a final time $T$. That is, we seek to maximize over choices of
harvest $\mathbf{E}_{X(t+1)} [ \sum_{t = 0}^{T}  R(X(t), h(t), t) \delta^t]$,
where the state dynamics are given by Eqn 1 and $\mathbf{E}$ denotes the
expectation over the process stochasticity in the future population state.

In order to find that policy, we introduce the value function $V(x(t), t)$
representing the total discounted catch from time $t$ onwards
given that $X(t) = x(t)$.  This value function satisfies an equation of
SDP  [@Mangel1988; @Clark2000; @Clark2009], sometimes called the
Hamilton-Jacobi-Bellman equation [@Mangel2014],

@@

where expectation is taken over all possible values of the next state,
$X(t+1)$, and maximized over all possible choices of harvest, $h(t)$.
That is, at time $t$, when population size is $x(t)$ and harvest $h(t)$ is
applied, the immediate return is $R(h(t), x(t))$. In the classical case,
the sole source of uncertainty is the process stochasticity term, $Z$,
and thus the expectation above is equivalent to taking expectations over
$Z(t)$. That is

\begin{equation}
\mathbf{\mathrm{E}}_{X(t+1)} \left[ V( X(t+1),t+1) | x(t), h(t) \right] = \mathbf{\mathrm{E}}_{Z(t)} \left[V( Z(t) f(x(t) - h(t))), t+1 | x(t), h(t) \right]
\end{equation}

where the population size after the take is $x(t) - h(t)$, which is then
translated into $X(t+1)$ by Eqn 1 (that is, we just replaced $X(t+1)$ with $Z(t) f(x(t)-h(t))$).

In a Bayesian decision framework, the parameters governing the dynamics
are also uncertain, so the expectation involves averaging over the
posterior distribution for the parameters, as well. That is,

\begin{equation}
\mathbf{\mathrm{E}}_{X(t+1)} \left[ V(X(t+1), t+1) | x(t), h(t) \right] = \mathbf{\mathrm{E}}_{p|\mathrm{data}} \{ \mathbf{\mathrm{E}}_{Z(t) | p, \mathrm{data}} \left[ V(Z(t) f(x(t) - h(t)), t)  \right] \}
\end{equation}

In the nonparametric case, the function $f$ too is uncertain and the expectation for the next state includes
uncertainty in $f$ as well. That is

\begin{equation}
\mathbf{\mathrm{E}}_{X(t+1)} \left[ V(X(t+1), t+1) | x(t), h(t) \right] = \mathbf{\mathrm{E}}_{p|\mathrm{data}} \{ \mathbf{\mathrm{E}}_{f, Z(t) | p, \mathrm{data}} \left[ V( Z(t) f(x(t) - h(t)), t) \right] \}
\end{equation}

@@

We consider three candidate parametric models of the stock-recruitment
dynamics: The Ricker model, the Allen model [@Allen2005], the Myers
model [@Myers1995], Eqns \eqref{ricker}-\eqref{myers}, which we parameterize as follows for $f(S(t),p)$ in Eqn 1. In all three, we let $K$ denote the carrying capacity and $r$ the maximum per capita growth rate. In the Ricker model,

\begin{equation}
f(S(t)|p) = S(t) e^{r \left(1 - \frac{S(t)}{K} \right) } \label{ricker}
\end{equation}

In the Allen model,

\begin{equation}
f(S(t)|p) = S(t) e^{r \left(1 - \frac{S(t)}{K}\right)\left(S(t) - X_c\right)} \label{allen}
\end{equation}

Where $r$ and $K$ are as before, and $X_c$ denotes the location of the unstable steady state (i.e., the tipping point).

In the Myers model,

\begin{equation}
f(S(t) | p)  = \frac{r S(t)^{\theta}}{1 + \frac{S(t)^\theta}{K}} \label{myers}
\end{equation}

Where $\theta = 1$ corresponding to Beverton-Holt dynamics and $\theta$ > 2 leads to Allee effects and multiple stable states.

The Ricker model involves two parameters, corresponding
to a growth rate and a carrying capacity, and cannot support alternative
stable growth rate dynamics. The Allen model resembles the Ricker
dynamics with an added Allee effect parameter [@Courchamp2008], below
which the population cannot persist.  The Myers model also has three
parameters and contains an Allee threshold, but has compensatory rather
than over-compensatory density dependence (resembling a Beverton-Holt
curve rather than a Ricker curve at high densities.)

The multiplicative log-normal stochasticity perturbs the growth predicted 
by each of these deterministic model skeletons. This 
introduces one additional parameter $\sigma$ that must be estimated.

Because of our interest in management performance in the presence of
tipping points, all of our simulations are based on the Allen model.
The Allen model is thus "structurally correct" and is expected to provide
a best-case scenario when the 'true' dynamics are known.
The Ricker model is a reasonable approximation of these dynamics far from the Allee
threshold (but lacks threshold dynamics), while the Myers model shares
the essential feature of a threshold but differs in the structure.

Throughout the presentation, we will refer to the use of the "True" model
to denote solving for the optimal solution when the underlying parameters
_are known without error_, and refer to the "Allen" model to denote the
when these parameters have instead only been estimated from the sample data.

We consider a period of `r Tobs` years of training data: long
enough that the estimates do not depend on the particular realization,
while longer times are not likely to provide substantial improvement.  
Each of the models is fit to the same training data (Figure 1).

@@

The core difference for our purpose between the GP representation and the parametric
models is that the GP model is defined explicitly in reference to 
some observed data.  Unlike the models above, it cannot be specified by
the value of some parameters alone.  Although called nonparametric, the
GP (a)  still involves the estimation of
parameters, and (b), statisticians also use non-parametric to mean models
that make only weak distributional assumptions  [e.g. @Lehmann1975].
However, we retain the terminology for consistency with previous studies.

@@

The use of GP methods to formulate a predictive
model is relatively new in the context of modeling dynamical systems
[@Kocijan2005], and was first introduced in the context ecological
modeling and fisheries management in @Munch2005.  GP models have
subsequently been used to test for the presence of Allee effects
[@Sugeno2013a], estimate the maximum reproductive rate [@Sugeno2013b],
determine temporal variation in food availability [@Sigourney2012], and
provide a basis for identifying model-misspecification [@Thorson2014].
An accessible and thorough introduction to the formulation and use of
GPs can be found in @Rasmussen2006.

A GP is an infinite dimensional (because it is a function of a continuous
input variable) stochastic process. Any realization of the process for
some finite sample of points $n$ is a multivariate normal distribution
of dimension $n$. Thus, to characterize the GP we need a mean function
and a covariance function. We proceed as follows.

We assume that the data $X(t)$ are observed with some process stochasticity,

@@

where $\varepsilon$ are IID normal random variables with zero-mean and variance $\sigma_n^2$. Note that we have chosen to
assume additive stochasticity.  While we could just as easily consider log-normal stochasticity
as in the parametric models, we make this choice to bring home the point that the
Gaussian process approach need not have structurally correct noise form either.

We consider the fase where we have some previous observations of the _timeseries_, $X_{\textrm{obs}}(t)$,
and wish to _predict_ future values $X_{\textrm{p}}(t)$.  We consider the observations as pairs of points
$X_{\textrm{obs}}(t), X_{\textrm{obs}}(t+1)$ for $t \in [1,2, ..., T]$ up through observed time $T$.  Under the GP, a finite sample of these predictions at $n$ points into the future, $X_{\textrm p}(\tau)$ for $\tau \in [T+1, T+2, ..., T+n]$, is simply given by a multivariate normal ($\mathcal{N}$):

@@

which has mean and covariance matrix given by:

@@

Where $K$ is the chosen covariance kernel.  (Compare to  Eqn 2.22 of @Rasmussen2006 for a more detailed derivation.) Many options are possible for this kernel. Throughout, we use the radial basis function for simplicity, a matrix whose $(i,j)$th element is given by:

\begin{equation}
K_{i,j}(x_i, y_j) = \exp\left(\frac{-|\mathbf{x_i}-\mathbf{j}|^2}{2 \ell^2} \right)
\end{equation}

We use inverse Gamma priors on both the length-scale $\ell$ and $\sigma$, of the form

\begin{equation}
f(x; \alpha, \beta) = \frac{\beta^\alpha}{\Gamma(\alpha)} x^{-\alpha - 1}\exp\left(-\frac{\beta}{x}\right)
\end{equation}

For the $\sigma$ prior, $\alpha = $ `r s2.p[1]` and $\beta = $ `r s2.p[2]`.  For $\ell$ prior,  $\alpha = $ `r d.p[1]` and $\beta = $ `r d.p[2]`.

Using the simulated training data we also estimate a Gaussian process
defined by a radial basis function kernel of two parameters: $\ell$,
which gives the characteristic length-scale over which correlation between
two points in state-space decays, and $\sigma$, which gives the scale
of the process noise by which observations $Y(t+1)$ may differ from
their predicted values $X(t+1)$ given an observation of the previous
state, $x(t)$ [see @Munch2005a].

@@

We use a Metropolis-Hastings Markov Chain Monte Carlo (@Gelman2003) to
infer posterior distributions of the two parameters of the GP (Figure
S3, code in appendix). Since the posterior distributions differ
substantially from the priors (Figure S3), we can be assured that most
of the information in the posterior comes from the data rather than the
prior belief.

@@

It is reasonably straight-forward to derive the harvest policy from the
estimated GP by inserting it into a SDP algorithm. The only difference 
is that the uncertainty in the future state under the GP, $X(t+1)$,
includes both process uncertainty (based on the estimation of $\sigma$) 
and "structural uncertainty" of the posterior collection of curves.  
(Recall that given the parameters, a parametric model is represented
by a single curve, while the GP remains a distribution of curves).  
From the GP posteriors, we can write down the (discritized) transition
matrix representing the probability of going to each state $X(t+1)$ 
given any current state $x(t)$ and any harvest $h(t)$ (See the function
`gp_transition_matrix()` in the provided R package). Given this transition
matrix, we use the same value iteration algorithm as in the parametric
case to determine the optimal policy.

@@

To ensure our results are robust to our choice of parameters, 
consider 96 different scenarios, described in detail below.   To help  better
understand the process, we first describe in detail the
results of a single scenario.

@@

All models fit the observed data rather closely and with relatively small uncertainty, as illustrated in the posterior predictive curves in Figure 1, in which we show the training data of stock sizes observed over time as points, overlaid with the step-ahead predictions of each estimated model using the parameters sampled from their posterior distributions.  Each model manages to fit the observed data rather closely. Compared to the true model most estimates appear to over-fit, predicting fluctuations that are actually due purely to stochasticity.  Model selection criteria (Table 1) penalize more complex models and show a preference for the simpler Ricker model over the  models with alternative stable states (Allen and Myers).  Details on MCMC estimates for each model, traces, and posterior distributions can be found in the appendix.

@@

```{r Table1, dependson="deviances", include=TRUE, results="asis", echo=FALSE}
pandoc.table(rbind(dictable, aictable, bictable), caption="Model selection scores for several common criteria (DIC: Deviance Information Criterion, AIC: Akaike Information Criterion, BIC: Bayesian Information Criterion) all (wrongly) select the simplest model. As the true (Allen) model is not distinguishable from the simpler (Ricker) model in the region of the observed data, this error cannot be avoided regardless of the model choice criterion. This highlights the danger of model choice when the selected model will be used outside of the observed range of the data.")
```

@@

write.csv(models_posteriors, "components/data/figure2.csv")
write.csv(obs, "components/data/obs.csv")
```

>>>>

Decision making under uncertainty is a ubiquitous challenge in the
management of human intervention in natural resources and conservation.
Decision-theoretic approaches provide a framework to determine the
best sequence of actions in face of uncertainty, but only when that
uncertainty can be meaningfully quantified [@Fischer2009].  Over the last
four decades (beginning with
@Clark1976, @Clark2009 and @Walters1978)
dynamic optimization methods, particularly Stochastic Dynamic Programming
(SDP), have become increasingly important as a means of
understanding how to manage human intervention into natural systems.
Simultaneously, there has been increasing recognition of the importance of
multiple steady states or `tipping points' [@Scheffer2001; .  @Scheffer2009;
@Polasky2011] in ecological systems.

We develop a novel approach to address these concerns in the context of
fisheries; although the challenges and methods are germane to
other problems of conservation or natural resource exploitation.  Economic value and ecological concern have made
marine fisheries the crucible for much of the founding work on management
under uncertainty [@Gordon1954;
@Clark1976; -@Clark2009; @Reed1979; @May1979; @Ludwig1982]

@@

Even if we know the proper deterministic description of the biological system, there is intrinsic stochasticity in biological dynamics, measurements, and
implementation of policy [_e.g._ @Reed1979; @Clark1986; @Roughgarden1996;
@Sethi2005].  We may also lack knowledge about the parameters of the biological dynamics [parametric uncertainty, _e.g._ @Ludwig1982; @Hilborn1997;
@McAllister1998; @Schapaugh2013], or even not know which model is  proper description of the system 
[structural uncertainty, _e.g._ @Williams2001; @Cressie2009;  @Athanassoglou2012].  Of these,
the latter is generally the hardest to quantify. Typical approaches
confront the data with a collection of models, assuming that the true dynamics (or reasonable approximation) is among the collection and then use model choice or model averaging to arrive at a conclusion 
[@Williams2001; @Athanassoglou2012; @Cressie2009].  Even setting aside
other concerns (see @Cressie2009), these
approaches are unable to describe uncertainty
outside the observed data range.

@@

Structural  uncertainty is particularly insidious when we try to predict
outside of the range of observed data [@Mangel2001] because we are extrapolating into unknown regions.  In management applications, this extrapolation uncertainty is particularly important since (a) management involves considering actions that may move the
system outside the range of observed behavior, and (b) the decision tools (optimal control theory, SDP) rely on both reasonable estimates of the expected
outcomes and on the weights given to those outcomes
[_e.g._ @Weitzman2013].   Thus characterizing uncertainty is as
important as characterizing the expected outcome.

Tipping points in ecological dynamics [@Scheffer2001;
@Polasky2011] highlight this problem because 
precise models are not available and data are
limited such as around  high
stock levels or an otherwise desirable state. With perfect information
, one would know just how far a system could
be pushed before crossing the tipping point, and management would be
simple.  But we face imperfect models and limited data and, with tipping points,  
even small errors can have
very large consequences, as we shall illustrate later.  Because intervention is often (but not always, see
@Hughes2013) too late after a tipping point has been crossed, management
is often concerned with avoiding 
tipping points before any data about them is available or the 
transition that would more clearly reveal a tipping point occurs
[e.g. @Bestelmeyer2012].

The dual concerns of model uncertainty and incomplete data 
create a substantial challenge to existing decision-theoretic approaches
[@Brozovic2011].   We illustrate how  Stochastic Dynamic Programming (SDP) 
[@Mangel1988; @Marescot2013] can be implemented using a
Bayesian Non-Parametric (BNP) model of population dynamics [@Munch2005a].
The BNP method has two distinct advantages 
.  First, using a BNP model sidesteps the need for an accurate
model-based description of the system dynamics.  Second, a BNP model
better reflects uncertainty when extrapolating beyond the observed
data.  This is crucial to providing  robust
decision-making when the correct model is not known (as is almost always true).
[We use _robust_ to characterize approaches that provide nearly optimal solutions
without being sensitive to the choice of the (unknown) underlying model.]

This paper is the first ecological application of the
SDP without an _a priori_  model of the
underlying dynamics.  Unlike parametric approaches that can only
reflect uncertainty in parameter estimates, the BNP method provides a
broader representation of uncertainty, including uncertainty beyond the observed data. 
We will show that Gaussian Process Dynamic Programming
(GPDP)  allows us to find robust management solutions in face of limited
data without knowing  the correct model structure.

@@

For comparisons, we consider the performance of management based on  GPDP against
management policies derived under several alternative parametric models [@Reed1979; @Ludwig1982; @Mangel1988].  Rather than compare models in terms
of best fit to data, we compare model performance in the concrete terms
of the decision-maker's objectives.

@@

[^1]: We abbreviate Gaussian Process as GP, which refers to 
the statistical model we use to approximate the population dynamics
, and we use the
term Gaussian Process Dynamic Programming [GPDP,], to refer to the _use_ of a
GP as the underlying process model when solving a Dynamic Programming equation
.  Hence we will refer to the models as: GP, Ricker, Allen, etc, and the novel method we put forward here as GPDP.

@@

Dynamic optimization  requires characterizing the dynamics of a state
variable (or variables), a control action, and a value function.
For simplicity, we consider only a single state variable, which allows us to
focus on one-dimensional models of population dynamics 
for the comparison of parametric models and the GP.
This is a best-case scenario for the parametric models because we
simulate underlying dynamics from one of the three parametric models
, whereas in the natural world we never know the "true" mode
.  In addition, by choosing one-dimensional models with just a few parameters, we
limit the chance that poor performance will be due to inability to
estimate parameters accurately, something that becomes a more severe problem
for higher-dimensional parametric models. Finally, the parametric models
we consider are those most commonly used in modeling stock-recruitment
dynamics or to model sudden transitions between alternative stable states.

@@

We let $X(t)$ denote the size (numbers or biomass) of the focal
population at time $t$ and assume that in the absence of take its
dynamics are.
.

\begin{equation}
X(t+1) = Z(t) f(X(t), p) \label{eq1} 
\end{equation}

Where Z(t) is log-normally distributed process stochasticity [@Reed1979]
and $p$ is a vector of parameters to be estimated from the data.
We describe the three choices for $f(X(t),p)$ in the next section.  
The control action is a harvest or take, $h(t)$, measured
in the same units as $X$, at time $t$.  Thus, in the presence of take,
the population size on the right hand side of Eqn 1 is replaced by
$S(t) = X(t) - h(t)$.

To construct the value function, we consider a return when $X(t) = x(t)$ and
harvest $h(t) = h$ denoted as the reward, $R(x(t), h)$. For example, if the
return is the harvest at time $t$, then $R(x(t), h(t)) = \min(x(t), h(t))$. We
assume that future harvests are discounted relative to current ones at
a constant rate of discount $\delta$ and ask for the harvest policy
that maximizes total discounted harvest between the current time $t$
and a final time $T$. That is, we seek to maximize over choices of
harvest $E_{X(t+1)} [ \sum_{t = 0}^{T}  R(X(t), h(t), t) \delta^t]$,
where the state dynamics are given by Eqn 1 and $E$ denotes the
expectation over the process stochasticity in the future population state.

In order to find that policy, we introduce the value function $V(x(t), t)$
representing the total discounted catch from time $t$ onwards
given that $X(t) = x(t)$.  This value function satisfies an equation of
SDP  [@Mangel1988; @Clark2000; @Clark2009 ;@Mangel2014],

@@

where expectation is taken over all possible values of the next state,
$X(t+1)$, and maximized over all possible choices of harvest, $h(t)$.
That is, at time $t$, when population size is $x(t)$ and harvest $h(t)$ is
applied, the immediate return is $R(h(t), x(t))$. When
the sole source of uncertainty is the process stochasticity term, $Z$,
and thus the expectation in Eqn 2 is equivalent to taking expectations over
$Z(t)$. That is

\begin{equation}
\mathbf{\mathrm{E}}_{X(t+1)} \left[ V( X(t+1),t+1) | x(t), h(t) \right] = \mathbf{\mathrm{E}}_{Z(t)} \left[V( Z(t) f(x(t) - h(t))|p), t+1 | x(t), h(t) \right]
\end{equation}

where the population size after the take is $x(t) - h(t)$, which is then
translated into $X(t+1)$ by Eqn 1 (that is, we replace $X(t+1)$ by $Z(t) f(x(t)-h(t)|p)$).

When the parameters governing the dynamics
are also uncertain (the case of parametric uncertainty), the expectation in Eqn 2 involves averaging over the
posterior distribution for the parameters. That is,

\begin{equation}
\mathbf{\mathrm{E}}_{X(t+1)} \left[ V(X(t+1), t+1) | x(t), h(t) \right] = \mathbf{\mathrm{E}}_{p|\mathrm{data}} \{ \mathbf{\mathrm{E}}_{Z(t) | p, \mathrm{data}} \left[ V(Z(t) f(x(t) - h(t)|p), t)  \right] \}
\end{equation}

When the underlying population dynamics are unknown (the case of structural uncertainty), the function $f$ itself is uncertain and the expectation for the next state includes
uncertainty in $f$ as well. That is

\begin{equation}
\mathbf{\mathrm{E}}_{X(t+1)} \left[ V(X(t+1), t+1) | x(t), h(t) \right] = \mathbf{\mathrm{E}}_{p|\mathrm{data}} \{ \mathbf{\mathrm{E}}_{f, Z(t) | p, \mathrm{data}} \left[ V( Z(t) f(x(t) - h(t)|p), t) \right] \}
\end{equation}

@@

We consider three candidate parametric models for the population dynamics: The Ricker model, the Allen model [@Allen2005], and the Myers
model [@Myers1995], Eqns \eqref{ricker}-\eqref{myers}. In all three, we let $K$ denote the carrying capacity and $r$ the maximum per capita growth rate. The Ricker model has two parameters and the right hand side of Eqn 1 is

\begin{equation}
f(S(t)|r,K) = S(t) e^{r \left(1 - \frac{S(t)}{K} \right) } \label{ricker}
\end{equation}

The Allen model has three parameters

\begin{equation}
f(S(t)|r, K, X_c) = S(t) e^{r \left(1 - \frac{S(t)}{K}\right)\left(S(t) - X_c\right)} \label{allen}
\end{equation}

where $X_c$ denotes the location of the unstable steady state (i.e., the tipping point).

The Myers model also has three parameters

\begin{equation}
f(S(t) | r, K, \theta)  = \frac{r S(t)^{\theta}}{1 + \frac{S(t)^\theta}{K}} \label{myers}
\end{equation}

where $\theta = 1$ corresponds to Beverton-Holt dynamics and $\theta$ > 2 leads to Allee effects and multiple stable states.

The Ricker model does not lead to multiple steady states.
The Allen model resembles the Ricker
dynamics with an added Allee effect parameter [@Courchamp2008], below
which the population cannot persist.  The Myers model also has three
parameters and contains an Allee threshold, but unlike the Ricker model
saturates  at high population size.  The multiplicative log-normal stochasticity in Eqn 1
introduces one additional parameter $\sigma$ that must be estimated.

Because of our interest in management performance in the presence of
tipping points, all of our simulations are based on the Allen model.
The Allen model is thus the  state of nature  and is expected to provide
the best-case scenario. The Ricker model is a reasonable approximation of these dynamics far from the Allee threshold (but lacks threshold dynamics), while the Myers model shares
the essential feature of a threshold but differs in structure from the Allen model.
Throughout, we refer to the "True" model when the underlying parameters
_are known without error_, and refer to the "Allen" model 
when these parameters have been estimated from the sample data.

We consider a period of `r Tobs` in which data are obtained to estimate the parameters or the GP. This is long
enough that the estimates do not depend on the particular realization,
and longer times are not likely to provide substantial improvement.  
We fit each of the models is fit to the same data (Figure 1).

@@

The core difference for our purpose between the GP and the parametric
models is that the GP model is defined explicitly by the observed data.  That is, it cannot be specified by the value of parameters alone.

@@

The use of GPs to characterize dynamical systems is relatively new
[@Kocijan2005], and was first introduced in the context ecological
modeling and fisheries management in @Munch2005.  GP models have
subsequently been used to test for the presence of Allee effects
[@Sugeno2013a], estimate the maximum reproductive rate [@Sugeno2013b],
determine temporal variation in food availability [@Sigourney2012], and
provide a basis for identifying model-misspecification [@Thorson2014].
An accessible and thorough introduction to the formulation and use of
GPs can be found in @Rasmussen2006.

A GP is a 
stochastic process for which any realization of for
a  finite sample of points $n$ is a multivariate normal distribution
of dimension $n$. Thus, to characterize the GP we need a mean function
and a covariance function. We proceed as follows.

As before, we assume that the data $X(t)$ are observed with process stochasticity around a mean function $ g(X(t))$

@@

where $\varepsilon$ are IID normal random variables with zero-mean and variance $\sigma_n^2$. Note that we have chosen to
assume additive stochasticity.  While we could consider log-normal stochasticity
as in the parametric models, we make this choice to emphasize that the
Gaussian process approach need not have structurally correct stochasticity to be effective.

Now consider the situation where we have a set of  previous observations of the _timeseries_, $X_{\textrm{obs}}(t)$,
and wish to _predict_ future values $X_{\textrm{p}}(t)$.  We consider the observations as pairs of points
$X_{\textrm{obs}}(t), X_{\textrm{obs}}(t+1)$ for $t=1, 2, …$ up through final observation time $T$.  As described above, in the  GP, a finite sample of these predictions at $n$ points into the future, $X_{\textrm p}(\tau)$ for $\tau \in [T+1, T+2, ..., T+n]$, is a multivariate normal ($\mathcal{N}$):

@@

with mean and covariance matrix given by (see Eqn 2.22 of @Rasmussen2006 for a more detailed derivation)

@@

Where $K$, called the covariance kernel, measures how much influence one observation has on another.  In this paper, for $K$ use a matrix whose $(i,j)$th element is given by:

\begin{equation}
K_{i,j}(x_i, y_j) = \exp\left( \frac{ -(x_i - y_j)^2}{2 \ell^2} \right)
\end{equation}
so that $\ell$ gives the characteristic length-scale over which correlation between
two observations decays.

We use inverse Gamma priors on both the length-scale $\ell$ and $\sigma$, thus for example

\begin{equation}
f(\ell; \alpha, \beta) = \frac{\beta^\alpha}{\Gamma(\alpha)} \ell^{-\alpha - 1}\exp\left(-\frac{\beta}{\ell}\right)
\end{equation}

For the prior on  $\ell$,  $\alpha = $ `r d.p[1]` and $\beta = $ `r d.p[2]`.  The prior  on $\sigma$,  $\alpha = $ `r s2.p[1]` and $\beta = $ `r s2.p[2]`.

@@

We use a Metropolis-Hastings Markov Chain Monte Carlo (@Gelman2003) to
infer posterior distributions of the parameters of the GP (Figure
S3, code in appendix). Since the posterior distributions differ
substantially from the priors (Figure S3), most of the information in the posterior comes from the data rather than the
prior belief.

@@

We  derive the harvest policy from the
estimated GP by inserting it into a SDP algorithm. Given the GP posteriors, we construct the transition
matrix representing the probability of going to each state $X(t+1)$ 
given any current state $x(t)$ and any harvest $h(t)$ (See the function
`gp_transition_matrix()` in the provided R package). Given this transition
matrix, we use the same value iteration algorithm as in the parametric
case to determine the optimal policy.   
In doing so, 
the uncertainty in the future state under the GP, $X(t+1)$,
includes both process uncertainty (based on the estimation of $\sigma$) 
and structural uncertainty of the posterior collection of curves.  
Unlike the case of a parametric model, which is a single curve given the parameters, 
the GP remains a distribution of curves.

@@

To ensure our results are robust to the choice of parameters, we
consider 96 different scenarios, described in detail below.   To help  better
understand the process, we first describe in detail the
results of a single scenario.

@@

All of the models fit the observed data rather closely and with relatively small uncertainty. In Figure 1, we show the posterior predictive curves.  The  training data of stock sizes observed over time are points, overlaid with the step-ahead predictions of each estimated model using the parameters sampled from their posterior distributions. Compared to the true model most estimates appear to over-fit, predicting patterns that are actually due purely to stochasticity.  Model selection criteria (Table 1) penalize more complex models and show a preference for the simpler Ricker model over the  models with alternative stable states (Allen and Myers).  Details on MCMC estimates for each model, traces, and posterior distributions can be found in the appendix.

@@

```{r Table1, dependson="deviances", include=TRUE, results="asis", echo=FALSE}
pandoc.table(rbind(dictable, aictable, bictable), caption="Model selection scores for several common criteria (DIC: Deviance Information Criterion, AIC: Akaike Information Criterion, BIC: Bayesian Information Criterion) all select the wrong model. As the true (Allen) model is not distinguishable from the simpler (Ricker) model in the region of the observed data, this error cannot be avoided regardless of the model choice criterion. This highlights the danger of model choice when the selected model will be used outside of the observed range of the data.")
```

@@

write.csv(models_posteriors, "components/data/figure2.csv")
write.csv(obs, "components/data/training_data.csv")
```
